---
title: "Getting started with messieR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with messieR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
set.seed(20260815)

#currently written as source("../messieR_core.R").
#once split out, library(messieR)
if (!exists("sim_var1")) source("../messieR_core.R")
```

## What this is for?

Simulation-based power analysis for a person-specific VAR(1) inherits whatever
you assume about the data-generating process, such as Gaussian innovations,
total data completeness, and one set of parameters for everyone. Since these
aren't representative of what experience sampling researchers may generally
observe in the real world, you get a sample-size recommendation that is not 
representative. What `messieR` lets you do is vary those assumptions one at a 
time and measure what, each could be worth in your scenario.

Note that this vignette runs entirely on simulated data, so it can build without 
a network connection. The last section shows the path to real archives, which 
does need you to be connected.

## The shortest useful example

Specify a transition matrix, generate a series, fit it, and get intervals from
all three procedures.

```{r basic}
phi <- matrix(c(0.5, 0.1,
                0.1, 0.5), 2, byrow = TRUE)

is_stationary(phi)                      #check if eigenvalues inside the circle?

Y   <- sim_var1(100, phi, sigma = diag(2))
fit <- fit_var1(Y)
round(fit$phi_hat, 3)
```

```{r basic-boot}
res <- boot_var1_all(Y, B = 199)
res[res$coef == "phi12", c("procedure", "est", "ci_lo", "ci_hi", "pval")]
```

Every row carries the same `est`, because the three procedures share one point
estimate and differ only in how they quantify uncertainty. When you display
display all of `res`, you can see that the `se` column is populated only for the 
Wald row. On the other hand, the bootstrap rows capture uncertainty through 
bias-corrected estimates (which is `NA` for Wald).

## A couple of conventions worth knowing before you trust output

### Transitions and VAR(1)

A VAR(1) is estimated from transitions, so `fit_var1()` lags the series first 
and drops incomplete pairs second. A pair thus enters the regression only when 
both endpoints are observed **and adjacent in time**. Noting how this order 
matters, deleting rows before lagging is going to splice non-adjacent 
observations into transitions that never happened. This means those false pairs 
carry weaker dependence than real ones, which pulls coefficients toward zero.

We can observe this with missingness:

```{r transitions}
Ym <- inject_mcar(Y, rate = 0.30)             #30% of variable 2 removed

rows_complete <- mean(stats::complete.cases(Ym))
pairs_ok <- {
  a <- Ym[-1, , drop = FALSE]; b <- Ym[-nrow(Ym), , drop = FALSE]
  mean(stats::complete.cases(a) & stats::complete.cases(b))
}
c(rows = round(rows_complete, 3),
  transitions = round(pairs_ok, 3),
  predicted = round(rows_complete^2, 3))
```

Seventy percent of rows leaves roughly half the transitions. A thirty percent
compliance gap costs about half your usable data, so if you're able to increase
compliance, more power to you.

### Coded gaps only

Deletion can exclude a gap only if the gap is recorded. Some public archives
store unanswered prompts as **absent rows** rather than `NA` rows; in such a 
case, the transition-level logic above cannot see anything to exclude. It's best 
to check this before fitting:

```{r gapcheck, eval = FALSE}
# if counter is the running prompt index for one participant...
#your_data <- ...
if(exists("your_data"))
  range(diff(sort(as.numeric(your_data$counter))))   
```

If that returns values over 1, reindex onto the prompt grid, inserting `NA`
rows for absent prompts and one `NA` row at each day boundary, so that no 
transition spans a night.


## Varying the innovation distribution

`sim_var1()` takes its innovations from one of three sources, in precedence: 
an empirical `residual_pool` beats a supplied `innov_fn`, which beats a Gaussian `sigma`.

```{r innov}
heavy <- function(n) matrix(rt(2 * n, df = 4), n, 2) / sqrt(4 / 2)

Y_gauss <- sim_var1(100, phi, sigma = diag(2))
Y_heavy <- sim_var1(100, phi, innov_fn = heavy)

rbind(gaussian = round(fit_var1(Y_gauss)$phi_hat[1, ], 3),
      heavy    = round(fit_var1(Y_heavy)$phi_hat[1, ], 3))
```

Consider how this substitution (say, skewed and heavy-tailed innovations) might
impact bias, coverage, and power, and whether this impact is outside Monte Carlo 
error of their Gaussian values. Consequently, it could be worth thinking on how
much effort to spend on getting the exact shape of an ESM variable's distribution,
versus compliance or assessing heterogeneity.

## Empirical innovations

A prospective use case of the `residual_pool` argument is to take a participant 
that you already have, keep their dynamics and their residual structure, and 
simulate the study you're about to run.

```{r pool}
mydat <- fit_var1(sim_var1(150, phi, innov_fn = heavy))   # stands in for real data

Y_new <- sim_var1(80, mydat$phi_hat,
                  delta = mydat$intercept,
                  residual_pool = mydat$residuals,
                  resample = "block")                     # keeps short-range structure
dim(Y_new)
```

Note that `resample = "block"` draws contiguous runs of residuals rather than
independent draws, with block length defaulting to
`default_block_length()`, the rule $\lceil 5.03\,T^{1/4}\rceil$.

## A small power analysis

Everything above composes into the question you actually care about: "how many
occasions do I need, given the compliance I expect?" So here's a demo that should
take a couple of minutes:

```{r power, cache = FALSE}
power_at <- function(T_use, rate, n_sim = 60L, B = 99L) {
  hits <- vapply(seq_len(n_sim), function(i) {
    Y  <- sim_var1(T_use, phi, sigma = diag(2))
    Yo <- if (rate > 0) inject_mcar(Y, rate) else Y
    r  <- tryCatch(boot_var1_all(Yo, B = B), error = function(e) NULL)
    if (is.null(r)) return(NA_integer_)
    r$reject_05[r$procedure == "residual_iid" & r$coef == "phi12"]
  }, integer(1))
  c(power = mean(hits, na.rm = TRUE),
    mcse  = mcse_proportion(hits))
}

round(rbind(`T=100, complete` = power_at(100, 0.00),
            `T=100, 30% MCAR` = power_at(100, 0.30),
            `T=150, 30% MCAR` = power_at(150, 0.30)), 3)
```

The settings here are small so that the vignette builds quickly, so
read the pattern rather than the digits. Report the Monte Carlo standard error 
alongside any number you take from a run like this. In fact at much smaller 
counts it is warranted, as the random noise might make, say, an MCAR case's power 
look slightly higher than the complete case purely by chance.

## Where this cannot help

While the bootstrap improves how the information you collected is used, it does
not add information. Missingness driven by the unrecorded state itself (i.e., 
"what if there's a reason or pattern to the holes in the data?") is a different 
problem, and nothing here addresses it.

The stationary bootstrap is included as an evaluated comparator. Resampling the 
observed series in blocks manufactures a transition at every block junction, and 
in the thesis it under-covered at all series lengths studied and over-declared 
effects on real data. `residual_iid` is preferable, unless you have a specific 
reason not to trust the VAR structure, in which case it could be worth comparing
to the residual-block variant of Brüggemann, Jentsch and Trenkler.

## Real archives

Datasets from the openESM database can be pulled directly. This chunk is not
evaluated, since building a vignette should not depend on a network.

```{r openesm, eval = FALSE}
library(openesm)

cat <- list_datasets()                     # metadata only
head(cat[, c("dataset_id", "first_author", "n_participants", "n_time_points")])

ds  <- get_dataset("0011")
dat <- as.data.frame(ds$data)

# inspect first: participant id, prompt index, and two continuous variables
str(dat)

# then, per participant, with ID_COL / VAR1 / VAR2 set from that inspection
Yp  <- cbind(as.numeric(dat[[VAR1]]), as.numeric(dat[[VAR2]]))[dat[[ID_COL]] == pid, ]
boot_var1_all(Yp, B = 999)
```

## Session

```{r session}
sessionInfo()
```
